### 2016-10-08 thesis_paper/source.tex
Last known good state
Last known good state goooood
automatic triggered changes
Formatting corrections

\documentclass[12pt,letterpaper]{report}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage[dvips]{graphicx}
\usepackage{xcolor}
\usepackage{times}
\usepackage{ragged2e}
\usepackage{titlesec}
\usepackage{wrapfig,subfigure,setspace,verbatim,cite,url,graphicx,float}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsfonts,amssymb,calc,calrsfs,cleveref}
\usepackage[justification=centering]{caption}
\usepackage{url}
\usepackage{subfigure}
\usepackage{cleveref}

@@

\usepackage{geometry}
\geometry{left=1in,right=1in,top=1in,bottom=1in}

\titleformat{\chapter}[hang] 
{\centering\normalfont\bfseries}{\MakeUppercase{\chaptertitlename}\ \thechapter:}{0.5em}{\uppercase}

\titleformat{\section}
  {\normalfont\fontsize{12}{15}\bfseries}{\thesection}{0.5em}{}

@@

\renewcommand{\contentsname}{TABLE OF CONTENTS}
\renewcommand{\listfigurename}{LIST OF FIGURES}
\renewcommand{\listtablename}{LIST OF TABLES}
\renewcommand{\abstractname}{ABSTRACT}

@@

Department of Computer Science and Engineering College of Engineering

@@

@@

August 23, 2016

@@

\linespread{1}\selectfont\clearpage
\thispagestyle{empty}
\centering
\linespread{1}\selectfont
\vspace*{3\baselineskip}
\textbf{DEDICATION}
\linespread{2}\selectfont
\begin{flushleft}
\hspace{3em} I dedicate my thesis work to my family who have always been supporting me throughout my life. I also dedicate it to my closest friends who have shared my passion and encouraged me to explore and try my ideas in life.
\end{flushleft}

@@

\linespread{1}\selectfont\clearpage
\thispagestyle{empty}
\centering
\linespread{1}\selectfont
\vspace*{3\baselineskip}
\textbf{ACKNOWLEDGMENTS}
\linespread{2}\selectfont
\begin{flushleft}
\hspace{3em} I would like to thank my major professor throughout my master thesis Dr. Yicheng Tu for his guidance and knowledge that I received through all the time I worked with him. I really appreciate his contribution and very happy that I got a chance to work with him. I as well want to thank the committee members for their support and encouragement.
\end{flushleft}

\clearpage
\restoregeometry

\linespread{1}\selectfont
\frontmatter
\tableofcontents

\linespread{2}\selectfont
\addcontentsline{toc}{chapter}{\listfigurename}
\listoffigures

\addcontentsline{toc}{chapter}{\listtablename} 
\listoftables

@@

\chapter*{Abstract}
\justify
\hspace{3em} Modern simulation systems generate big amount of data, which consequently has to be analyzed in a timely fashion. Traditional database management systems follow principle of pulling the needed data, processing it, and then returning the results. This approach is then optimized by means of caching, storing in different structures, or doing some sacrifices on precision of the results to make it faster. When it comes to the point of doing various queries that require analysis  of the whole data,  this design has the following disadvantages: considerable overhead on traditional disk random I/O framework while reading from the simulation output files and low throughput of the data that consequently results in long latency, and, if there was any indexing to optimize selections, overhead of storing those becomes too big, too. Besides it, indexing will also cause delay during write operations and since most of the queries work with the entire data sets, indexing loses its point.

There is a new approach to this problem -- Push-based System for Molecular Simulation Data Analysis for processing network of queries proposed in the previous paper and its primary steps are: i) it uses traditional scan-based I/O framework to load the data from files to the main memory and then ii) the data is pushed through a network of queries which consequently filters the data and collect all the needed information which increases efficiency and data throughput. It has a considerable advantage in analysis of molecular simulation data, because it normally involves all the data sets to be processed by the queries.

In this paper, we propose improved version of Push-based System for Molecular Simulation Data Analysis. Its major difference with the previous design is usage of GPU for the actual processing part of the data flow. Using the same scan-based I/O framework the data is pushed through the network of queries which are processed by GPU, and due to the nature of science simulation data, this gives a big advantage for processing it faster and easier (it will be explained more in later sections). In the old approach there were some custom data structures such as quad-tree for calculation of histograms to make the processing faster and those involved loss of data and some expectations from the data nature, too. In the new approach due to high performance of GPU processing and its nature, custom data structures were not even needed much, though it didn't bear any loss in precision and performance.

%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%
%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%

\mainmatter
\linespread{1}\selectfont
\chapter{Introduction} \linespread{2}\selectfont \justify

\hspace{3em} In various sciences simulation systems take big place and often times they may be the clue for results. One of such sciences, which is primarily related to this paper, is physics. In this case, from computer engineering point of view, simulations have the following flow:

@@

The purpose of this paper is primarily focused on the third step, which basically involves the entire simulation data to be processed by the analysis software. It is important to underline scale of data being discussed. Molecular simulations represent systems starting from hundreds thousands of atoms and more which are tracked by frames as snapshot of specific time and can last for months. Thus, the size of data starts from couple of gigabytes up to terabytes and petabytes.

Big data processing is becoming one of the key issues with the amount of data being generated by modern systems. Eventually, this data is required to be processed on the fly, thus, analysis software should be able to handle massive data in a very short period of time. When working with huge volume of data, there are non-trivial issues arise. For example, it can take days and weeks to analyze big enough data sets, because the data cannot be simply loaded into memory, since it can get up to terabytes, thus, it has extra overhead because of traditional random access disk I/O framework to read from disk chunks by chunks. Besides it, analysis of the data can get even more complicated, since going through certain parts of the data once would not be enough, which leads to low throughput and efficiency of loaded data. One such example, the data analysis approach has polynomial complexity just for reading the data in order to come up with result, and since the data can't simply be saved in memory, it raises the overhead of disk I/O, too. Pull-based architectures in data processing engines are inefficient, since having a set of specific queries, in order to compute them all, it is needed to fetch data, filter it, and apply needed formulas. It is inefficient, since for every query the same chunk of data needs to be pulled into the memory at least the number of queries times or more in case of more complex queries.

As it has already been mentioned in the previous paper \cite{mainPaper}, one of the modern issues of analyzing massive data on the fly is social networks. "In order for a system to be able to perform analytical examination of the data produced in such streaming media, the system should have the capability of fast data access. The reason is millions of data records (tweets) produced every second. Moreover, these tweets may have different geographical origin, introducing different languages and forms and often times containing unsolicited messages, errors,
malicious content, etc. Therefore, some low level data uniformity
and cleaning on top of the data access and management issues should
be considered and possibly incorporated in the process of analytic
investigation in order to achieve relevant result".\cite{mainPaper, nature_bigdata08,nature_bigdata12,science_social10,jcs_twitter11} Another good example would be data scale of Facebook Inc. Being the largest social network, it has huge data coming in daily \cite{fbStats}. The data warehouse of Facebook is measured to be 300PB in 2014, and the daily growth is 600TB, most of the raw data is logs, which, most of the times, are used for analytics. Analytical queries involve entire data sets that will be handy for any future strategical planning in big companies, being able to observe data behavior and much more.

The primary focus and problem in this paper is scientific data analysis. Examples brought up above are real life problems that are very similar to the scientific problem that we are focusing on in terms of data sizes. Particles simulation is one of the most popular methods of analyzing certain chemical reactions, physical processes, or other behavior of different materials.
Molecular simulations (Molecular Dynamics) are applied in different fields and represent a method of analyzing physical movements of particles, atoms, and molecules in a fixed space with a given period of time, apparently with a possibility of giving initial state for each item that is involved in the process and can affect the system. These systems are most of the time N-body simulations. The number of atoms in simulations vary in hundred of thousands, particularly, we may observe two simulation systems of a collagen fiber structure and dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system consisting of 890,000 and 402,400 atoms as mentioned and illustrated in previous work of the system \cite{mainPaper}. Simulation data represents number of records of physical properties such as mass, charge, velocity, coordinates, and forces for each item aggregated as frames, where each frame represents a snapshot of time, placed with a fixed time interval which may also vary depending on the simulation itself and simulation precision requirement. "Quantities measured during the simulations are analyzed to test the theoretical
model~\cite{Frenkel:api01,Landau:cup05}. In short, the MS is proven
and powerful tool for understanding the inner-workings of a
biological system, by supplying a model description of the
biophysical and biochemical processes that are being unfold at a
nanoscopic scale." \cite{mainPaper}

Scientist gives the properties to simulation software (for example, Gromacs), runs the simulations, and finally get the output file. The output file must then be analyzed to produce certain results which may help him come up with certain consensus on original theoretical model that resulted in the molecular simulation system\cite{Frenkel:api01}. Gromacs is simulation software tool that helps scientist to run the actual simulation. It is a molecular dynamics package primarily designed for biomolecular systems such as proteins and lipids. \cite{Gromacs-online}. Besides the fact that it helps to generate the output files for the simulations, apparently it also helps to analyze the data itself, but the original problem is that it is not as optimized as it can be in order to analyze the data. Gromacs follows approach of pull-based design, which means that for any given query (e.g. total mass or total charge, which are very similar type of 1-body queries without sophisticated selection) it will pull data separately and generate addition overhead wasting disk read I/O in order to come up with the result just for a single query. As it has been proposed in the previous paper, in order to remedy such issues, the push-based design does exactly the opposite, where instead of loading the data on demand for each query, the queries are batched into a network, then the entire dataset is loaded chunk by chunk pushing it through the network which has its internal relationships and dependencies amonth the queries. In this case, since scientific simulation data is run once with specific physical properties, it is never modified, thus, on continuation, it will only append, which means that the processing can also be run on appended frames. \cite{mainPaper}
This type of approach has already been revised by other systems \cite {DataPath,Volcano,Qpipe} as of reusing loaded data through queries produced \cite {Candea,PredictablePerformance,CooperativeScans}.

In this paper, we are incorporating parallelization into the processing part of the proposed design by means of CUDA programming language for GPU. Since storage of the simulation data is very expensive, it might come to the point of analyzing the data on the fly (meaning running the simulation and analyzing it at the same time in a streaming manner), which leads to a problem of optimizing the processing part of the design proposed in the previous paper, because time spent on generating data should be tried to conceal the time spent on the processing part by means of overlapping or simply running it in a quick manner. We will talk about this in details as well in further sections.

\section{Problem Statement}
\hspace{3em} Simulation software systems, in general, follow the same methodology of running and storing simulation data. The simulation software system examples are: Gromacs \cite{GROMACS4}, VMD\cite{VMD}, MDAnalysis~\cite{MDAnalysis}, Wordom~\cite{wordom}, MD-TRACKS~\cite{MDtracks}, SimulaidOne~\cite{Simulaid}, Charmm~\cite{CHARMM}. In the type of simulations brought up as examples above, the flow of the data is the following. Once the simulation is run, the output files are contained as trajectory files with descriptors (they contain information about space dimensions, number of atoms and frames, etc.) that can be easily transposed into simple flat files containing the physical properties atom by atom, frame by frame, which are consequently read and processed by the proposed push-based system. Since we have certain amount of queries needed to be run on given simulation data, generating high I/O traffic followed by design of pull-based system is not considerable. The approach proposed in the previous paper is very good in terms of performance in comparison with the original pull-based system\cite{mainPaper}. As you can see on Figure \ref{fg:sample_estimation_old_paper}, it gets up to over 100 times of speedup which is quite crucial, if we think about days of computations. The problem is still that some of the queries processed by those means are still improvable, especially taking into consideration the fact that in the used previous works for calculation of 2-body functions, which take the biggest time for processing~\cite{ytu:icde09, EDBT12}, we might have some error bounds, which might be unacceptable for certain simulation analysis where sacrifice on loss of data is unbearable. Besides it, although we might have huge performance boost on specific data structure as density map, we still have certain expectation from data nature (such as its uniformity), thus, it makes sense to have desire to simplify the processing and still having pure computation based on any values, considering that the queries should be available to be pre-programmed by user in a separate module of the code. All the benchmarks will be done in comparison with already push-based system versus push-based system with GPU improvements. We think that it is  not fair to compare pull-based against GPU push-based approach, since the goal of the paper is to demonstrate capabilities of GPU approach that can beat simple sequential push-based system.

\section{Our approach with improvement}

@@

Original non-improvement standard approach with pull-based querying is probably one of the most flexible, because nobody has to think about improvements and complexities, it does it for every chunk of data and pulls it whenever it is needed. As it has already been mentioned, at this point, the biggest challenge is that supposing we have $M$ queries, for $N$ atoms we would have to load every chunk $M$ times for $N$ chunks of data (frames), and apparently it is very time consuming of I/O and non-linear computations, and that is only for linear functions. Improvements could be done on caching level, but again the worst case would be the same complexity, because most of the time there is nothing really to cache, and if there is, it is too big to store it in memory, so, it just has to be re-read from file. Using accumulative data structures and trying to optimize by means of them doesn't help either, because, first of all, it is too big data to store in memory, and, secondly, most of the time it is based on data distribution expectation of some kind, which is not the best solution, because the experiments can be different, thus data distribution is sometimes not predictable (boiling down to a specific physics problem will destroy the abstractness and applicability of the system to variety of other problems and will narrow it down making it less useful). In summary, the bottlenecks in this approach are I/O time consumption and slow data processing (especially for non-linear computation).

After all, push-based system for solution of the given problem is tremendously fast in comparison with pull-based system. General idea is to push a chunk of data (frame) through the pre-defined network of queries that will be needed to run on the data sets, which basically means chunk of data will be read from file and loaded into memory only once in order to go through all the queries. As you can observe \cite{mainPaper}, it can speed up for more than 100 times (depending on workload, number of atoms, and functions being used). It is clear that the biggest challenges in this approach were related to implementation of non-linear functions like SDH, since once the data is pushed through the queries, it is almost not coming back (if it is coming back, the point of optimization is lost), and the idea is to come up with approach to eliminate this overhead and weakness. In this case, it is also important what memory exactly is being used, because trying not to use used parts of the same chunk also matters, because L-level caches of CPU might play a huge role in big data sets. One of the solutions is building quad-tree based data structure that will help optimize the calculation, and it has some expectation from data distribution which can speed up depending on dimensions (2D and 3D). This is a mutual problem of both data loading (memory usage) and its processing, and push-based system is a great approach, because it still remains as an abstract concept applicable to different queries and probably wide variety of other problems of similar kind in different fields.

Finally, we've analyzed some main ideas, advantages, and disadvantages of our approach in this paper which is basically push-based system with optimization on data processing part by means of GPU computation (in other words, parallelism). Reading the data from files is a huge overhead, thus, speedup of push-based system in general is significant over pull-based system in our given case. 
It's been demonstrated in the previous paper with different amount of atoms, frames, and workloads. Since we have to load the data every time there is a different query versus push-based system loads a chunk of data once, it is quite noticeable how the framework contributes to efficiency of analysis of simulation data. Even though the memory loading and pushing steps are the key features of the proposed design, nevertheless, having general knowledge about the network of queries, in this paper, we will try to focus on optimizing the actual execution of them.

Some of the queries may be very slow due to their nature on a sequential type of computation,  especially, 2-body functions. Since the nature of the data that comes with simulation is basically physical properties of atoms, there is a lot of computation that involves independent primitive mathematical operations, which makes it a perfect problem for parallelism. For example, image processing computations and graphics processing are done on GPUs, because they have a lot of minor primitive independent pixel-based computations.

As it has already been mentioned, although the original idea of push-based system for molecular simulation data analysis was focused on optimizing throughput and usage of loaded data, there were some extra approaches specifically for 2-body functions. For example, Density Map for Spacial Distance Histogram was used in order to avoid additional memory allocation and latency reduction with proven error bounds. SDH is a quite intensive and computational problem, especially with increasing amount of atoms.

We believe that having this nature of computational problems, the proposed GPU improved version of push-based system will significantly change in terms of performance by incorporating parallelism with CUDA.

@@

\section{Contribution and roadmap of the paper}

@@

\chapter{Related Work}\label{sc:relatedwork}

@@

\section{Scientific software}

@@

The queries that we consider are related to computations limited within one frame of a molecular simulation. In general, with the given queries we have divided them into two categories: one body functions and multiple body functions. The first ones apparently have linear complexity of $O(n)$, while the other ones have bigger complexity.

\section{One body functions}
\hspace{3em} \emph{\textbf{One body functions}} such as sum of masses, center of mass, or simple counting with selection  are of a complexity $O(n)$, thus, in order to get the results for them going through every atom (essentially over the entire data) only once is enough, and taking into consideration the fact that we are going with push-based system, it is done at the same time for all of them.

>>>>

\documentclass[12pt,letterpaper]{report}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage[dvips]{graphicx}
\usepackage{xcolor}
\usepackage{times}
\usepackage{ragged2e}
\usepackage{titlesec}
\usepackage{wrapfig,setspace,verbatim,cite,url,graphicx,float}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsfonts,amssymb,calc,calrsfs,cleveref}
\usepackage[justification=centering]{caption}
\usepackage{url}
\usepackage{cleveref}

@@

\usepackage{geometry}
\geometry{left=1in,right=1in,top=1in,bottom=1.2in}

\titleformat{\chapter}[hang]{\normalfont\normalsize\filcenter\bfseries}{\centering\MakeUppercase{\chaptertitlename}\ \thechapter:}{0.5em}{\uppercase}

\titleformat{\section}
  {\normalfont\fontsize{12}{15}\bfseries}{\thesection}{0.5em}{}
\titleformat{\subsection}
  {\normalfont\fontsize{12}{15}\bfseries}{\thesection}{0.5em}{} \titleformat{\subsubsection}
  {\normalfont\fontsize{12}{15}\bfseries}{\thesection}{0.5em}{}
%toc

\titlespacing*{\chapter} {0pt}{50pt}{15pt}
\titlespacing*{\section}{0pt}{0\baselineskip}{0\baselineskip}
\titlespacing*{\subsection}{0pt}{0\baselineskip}{0\baselineskip}
\titlespacing*{\subsubsection}{0pt}{0\baselineskip}{0\baselineskip}

@@

\renewcommand{\contentsname}{\normalsize TABLE OF CONTENTS}
\renewcommand{\listfigurename}{\normalsize LIST OF FIGURES}
\renewcommand{\listtablename}{\normalsize LIST OF TABLES}
\renewcommand{\abstractname}{\normalsize ABSTRACT}
\renewcommand{\bibname}{\normalsize REFERENCES}

\usepackage{tocloft}
\renewcommand{\cftchapleader}{\cftdotfill{\cftdotsep}}
\renewcommand{\cftpartleader}{\cftdotfill{\cftdotsep}} 
\renewcommand{\cftchapfont}{\normalfont}
\renewcommand{\cftchappagefont}{\MakeUppercase\normalfont}
\newlength\mylength
\renewcommand\cftchappresnum{\MakeUppercase{\chaptername~}}
\renewcommand\cftchapaftersnum{:}

\setlength\cftaftertoctitleskip{20pt}
\setlength\cftafterlottitleskip{15pt}
\setlength\cftafterloftitleskip{15pt}

\newcommand*{\noaddvspace}{\renewcommand*{\addvspace}[1]{}}
\addtocontents{lof}{\protect\noaddvspace}
\addtocontents{lot}{\protect\noaddvspace}

\renewcommand{\cftfigpresnum}{Figure\ }
\renewcommand{\cfttabpresnum}{Table\ }

\cftsetindents{section}{0.5in}{0.3in}
\cftsetindents{subsection}{1in}{0.4in}
\cftsetindents{subsubsection}{1.5in}{0.5in}

\setlength{\cfttabindent}{0in}
\setlength{\cftfigindent}{0in}

\newlength{\mylenf}
\settowidth{\mylenf}{\cftfigpresnum}
\setlength{\cftfignumwidth}{\dimexpr\mylenf+2em}
\setlength{\cfttabnumwidth}{\dimexpr\mylenf+2em}

\settowidth\mylength{\cftchappresnum\cftchapaftersnum}
\addtolength\cftchapnumwidth{\mylength-0.5em}

@@

Department of Computer Science and Engineering

College of Engineering

@@

Srinivas Katkoori, Ph. D.

Sameer Varma, Ph. D.

@@

October 10, 2016

@@

\linespread{1}\selectfont\clearpage
\thispagestyle{empty}
\centering
\linespread{1}\selectfont
\vspace*{3\baselineskip}
\textbf{DEDICATION}
\linespread{2}\selectfont
\justify
\hspace{3em} I dedicate my thesis work to my family who have always been supporting me throughout my life. I also dedicate it to my closest friends who have shared my passion and encouraged me to explore and try my ideas in life.

@@

\linespread{1}\selectfont\clearpage
\thispagestyle{empty}
\centering
\linespread{1}\selectfont
\vspace*{3\baselineskip}
\textbf{ACKNOWLEDGMENTS}
\linespread{2}\selectfont
\justify
\hspace{3em} I would like to thank my major professor throughout my master thesis Dr. Yicheng Tu for his guidance and knowledge that I received through all the time I worked with him. I really appreciate his contribution and very happy that I got a chance to work with him. I as well want to thank the committee members for their support and encouragement.

\linespread{1}\selectfont
\vspace{1in}

\frontmatter
\centering
\tableofcontents

\clearpage
\linespread{2}\selectfont
\addcontentsline{toc}{chapter}{\listtablename} 
\listoftables

\clearpage
\addcontentsline{toc}{chapter}{\listfigurename}
\listoffigures

@@

\chapter*{Abstract}
\justify
\hspace{3em}Modern simulation systems generate big amount of data, which consequently has to be analyzed in a timely fashion. Traditional database management systems follow principle of pulling the needed data, processing it, and then returning the results. This approach is then optimized by means of caching, storing in different structures, or doing some sacrifices on precision of the results to make it faster. When it comes to the point of doing various queries that require analysis  of the whole data,  this design has the following disadvantages: considerable overhead on traditional disk random I/O framework while reading from the simulation output files and low throughput of the data that consequently results in long latency, and, if there was any indexing to optimize selections, overhead of storing those becomes too big, too. Beside it, indexing will also cause delay during write operations and since most of the queries work with the entire data sets, indexing loses its point.

\noindent\hspace{3em}There is a new approach to this problem -- Push-based System for Molecular Simulation Data Analysis for processing network of queries proposed in the previous paper and its primary steps are: i) it uses traditional scan-based I/O framework to load the data from files to the main memory and then ii) the data is pushed through a network of queries which consequently filters the data and collect all the needed information which increases efficiency and data throughput. It has a considerable advantage in analysis of molecular simulation data, because it normally involves all the data sets to be processed by the queries.

\noindent\hspace{3em}In this paper, we propose improved version of Push-based System for Molecular Simulation Data Analysis. Its major difference with the previous design is usage of GPU for the actual processing part of the data flow. Using the same scan-based I/O framework the data is pushed through the network of queries which are processed by GPU, and due to the nature of science simulation data, this gives a big advantage for processing it faster and easier (it will be explained more in later sections). In the old approach there were some custom data structures such as quad-tree for calculation of histograms to make the processing faster and those involved loss of data and some expectations from the data nature, too. In the new approach due to high performance of GPU processing and its nature, custom data structures were not even needed much, though it didn't bear any loss in precision and performance.

\mainmatter
\linespread{1}\selectfont
\chapter{INTRODUCTION} 
\linespread{2}\selectfont\justify
\hspace{3em} In various sciences simulation systems take big place and often times they may be the clue for results. One of such sciences, which is primarily related to this paper, is physics. In this case, from computer engineering point of view, simulations have the following flow:

@@

\noindent\hspace{3em}The purpose of this paper is primarily focused on the third step, which basically involves the entire simulation data to be processed by the analysis software. It is important to underline scale of data being discussed. Molecular simulations represent systems starting from hundreds thousands of atoms and more which are tracked by frames as snapshot of specific time and can last for months. Thus, the size of data starts from couple of gigabytes up to terabytes and petabytes.

\noindent\hspace{3em}Big data processing is becoming one of the key issues with the amount of data being generated by modern systems. Eventually, this data is required to be processed on the fly, thus, analysis software should be able to handle massive data in a very short period of time. When working with huge volume of data, there are non-trivial issues arise. For example, it can take days and weeks to analyze big enough data sets, because the data cannot be simply loaded into memory, since it can get up to terabytes, thus, it has extra overhead because of traditional random access disk I/O framework to read from disk chunks by chunks. Beside it, analysis of the data can get even more complicated, since going through certain parts of the data once would not be enough, which leads to low throughput and efficiency of loaded data. One such example, the data analysis approach has polynomial complexity just for reading the data in order to come up with result, and since the data can't simply be saved in memory, it raises the overhead of disk I/O, too. Pull-based architectures in data processing engines are inefficient, since having a set of specific queries, in order to compute them all, it is needed to fetch data, filter it, and apply needed formulas. It is inefficient, since for every query the same chunk of data needs to be pulled into the memory at least the number of queries times or more in case of more complex queries.

\noindent\hspace{3em}As it has already been mentioned in the previous paper \cite{mainPaper}, one of the modern issues of analyzing massive data on the fly is social networks. "In order for a system to be able to perform analytical examination of the data produced in such streaming media, the system should have the capability of fast data access. The reason is millions of data records (tweets) produced every second. Moreover, these tweets may have different geographical origin, introducing different languages and forms and often times containing unsolicited messages, errors,
malicious content, etc. Therefore, some low level data uniformity
and cleaning on top of the data access and management issues should
be considered and possibly incorporated in the process of analytic
investigation in order to achieve relevant result".\cite{mainPaper, nature_bigdata08,nature_bigdata12,science_social10,jcs_twitter11} Another good example would be data scale of Facebook Inc. Being the largest social network, it has huge data coming in daily \cite{fbStats}. The data warehouse of Facebook is measured to be 300PB in 2014, and the daily growth is 600TB, most of the raw data is logs, which, most of the times, are used for analytics. Analytical queries involve entire data sets that will be handy for any future strategical planning in big companies, being able to observe data behavior and much more.

\noindent\hspace{3em}The primary focus and problem in this paper is scientific data analysis. Examples brought up above are real life problems that are very similar to the scientific problem that we are focusing on in terms of data sizes. Particles simulation is one of the most popular methods of analyzing certain chemical reactions, physical processes, or other behavior of different materials.
Molecular simulations (Molecular Dynamics) are applied in different fields and represent a method of analyzing physical movements of particles, atoms, and molecules in a fixed space with a given period of time, apparently with a possibility of giving initial state for each item that is involved in the process and can affect the system. These systems are most of the time N-body simulations. The number of atoms in simulations vary in hundred of thousands, particularly, we may observe two simulation systems of a collagen fiber structure and dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system consisting of 890,000 and 402,400 atoms as mentioned and illustrated in previous work of the system \cite{mainPaper}. Simulation data represents number of records of physical properties such as mass, charge, velocity, coordinates, and forces for each item aggregated as frames, where each frame represents a snapshot of time, placed with a fixed time interval which may also vary depending on the simulation itself and simulation precision requirement. "Quantities measured during the simulations are analyzed to test the theoretical
model~\cite{Frenkel:api01,Landau:cup05}. In short, the MS is proven
and powerful tool for understanding the inner-workings of a
biological system, by supplying a model description of the
biophysical and biochemical processes that are being unfold at a
nanoscopic scale." \cite{mainPaper}

\noindent\hspace{3em}Scientist gives the properties to simulation software (for example, Gromacs), runs the simulations, and finally get the output file. The output file must then be analyzed to produce certain results which may help him come up with certain consensus on original theoretical model that resulted in the molecular simulation system\cite{Frenkel:api01}. Gromacs is simulation software tool that helps scientist to run the actual simulation. It is a molecular dynamics package primarily designed for biomolecular systems such as proteins and lipids. \cite{Gromacs-online}. Besides the fact that it helps to generate the output files for the simulations, apparently it also helps to analyze the data itself, but the original problem is that it is not as optimized as it can be in order to analyze the data. Gromacs follows approach of pull-based design, which means that for any given query (e.g. total mass or total charge, which are very similar type of 1-body queries without sophisticated selection) it will pull data separately and generate addition overhead wasting disk read I/O in order to come up with the result just for a single query. As it has been proposed in the previous paper, in order to remedy such issues, the push-based design does exactly the opposite, where instead of loading the data on demand for each query, the queries are batched into a network, then the entire dataset is loaded chunk by chunk pushing it through the network which has its internal relationships and dependencies amonth the queries. In this case, since scientific simulation data is run once with specific physical properties, it is never modified, thus, on continuation, it will only append, which means that the processing can also be run on appended frames. \cite{mainPaper}
This type of approach has already been revised by other systems \cite {DataPath,Volcano,Qpipe} as of reusing loaded data through queries produced \cite {Candea,PredictablePerformance,CooperativeScans}.

\noindent\hspace{3em}In this paper, we are incorporating parallelization into the processing part of the proposed design by means of CUDA programming language for GPU. Since storage of the simulation data is very expensive, it might come to the point of analyzing the data on the fly (meaning running the simulation and analyzing it at the same time in a streaming manner), which leads to a problem of optimizing the processing part of the design proposed in the previous paper, because time spent on generating data should be tried to conceal the time spent on the processing part by means of overlapping or simply running it in a quick manner. We will talk about this in details as well in further sections.

\section{Problem Statement}
\hspace{3em} Simulation software systems, in general, follow the same methodology of running and storing simulation data. The simulation software system examples are: Gromacs \cite{GROMACS4}, VMD\cite{VMD}, MDAnalysis~\cite{MDAnalysis}, Wordom~\cite{wordom}, MD-TRACKS~\cite{MDtracks}, SimulaidOne~\cite{Simulaid}, Charmm~\cite{CHARMM}. In the type of simulations brought up as examples above, the flow of the data is the following. Once the simulation is run, the output files are contained as trajectory files with descriptors (they contain information about space dimensions, number of atoms and frames, etc.) that can be easily transposed into simple flat files containing the physical properties atom by atom, frame by frame, which are consequently read and processed by the proposed push-based system. Since we have certain amount of queries needed to be run on given simulation data, generating high I/O traffic followed by design of pull-based system is not considerable. The approach proposed in the previous paper is very good in terms of performance in comparison with the original pull-based system\cite{mainPaper}. As you can see on Figure \ref{fg:sample_estimation_old_paper}, it gets up to over 100 times of speedup which is quite crucial, if we think about days of computations. The problem is still that some of the queries processed by those means are still improvable, especially taking into consideration the fact that in the used previous works for calculation of 2-body functions, which take the biggest time for processing~\cite{ytu:icde09, EDBT12}, we might have some error bounds, which might be unacceptable for certain simulation analysis where sacrifice on loss of data is unbearable. Beside it, although we might have huge performance boost on specific data structure as density map, we still have certain expectation from data nature (such as its uniformity), thus, it makes sense to have desire to simplify the processing and still having pure computation based on any values, considering that the queries should be available to be pre-programmed by user in a separate module of the code. All the benchmarks will be done in comparison with already push-based system versus push-based system with GPU improvements. We think that it is  not fair to compare pull-based against GPU push-based approach, since the goal of the paper is to demonstrate capabilities of GPU approach that can beat simple sequential push-based system.

\section{Our Approach With Improvement}

@@

\noindent\hspace{3em}Original non-improvement standard approach with pull-based querying is probably one of the most flexible, because nobody has to think about improvements and complexities, it does it for every chunk of data and pulls it whenever it is needed. As it has already been mentioned, at this point, the biggest challenge is that supposing we have $M$ queries, for $N$ atoms we would have to load every chunk $M$ times for $N$ chunks of data (frames), and apparently it is very time consuming of I/O and non-linear computations, and that is only for linear functions. Improvements could be done on caching level, but again the worst case would be the same complexity, because most of the time there is nothing really to cache, and if there is, it is too big to store it in memory, so, it just has to be re-read from file. Using accumulative data structures and trying to optimize by means of them doesn't help either, because, first of all, it is too big data to store in memory, and, secondly, most of the time it is based on data distribution expectation of some kind, which is not the best solution, because the experiments can be different, thus data distribution is sometimes not predictable (boiling down to a specific physics problem will destroy the abstractness and applicability of the system to variety of other problems and will narrow it down making it less useful). In summary, the bottlenecks in this approach are I/O time consumption and slow data processing (especially for non-linear computation).

\noindent\hspace{3em}After all, push-based system for solution of the given problem is tremendously fast in comparison with pull-based system. General idea is to push a chunk of data (frame) through the pre-defined network of queries that will be needed to run on the data sets, which basically means chunk of data will be read from file and loaded into memory only once in order to go through all the queries. As you can observe \cite{mainPaper}, it can speed up for more than 100 times (depending on workload, number of atoms, and functions being used). It is clear that the biggest challenges in this approach were related to implementation of non-linear functions like SDH, since once the data is pushed through the queries, it is almost not coming back (if it is coming back, the point of optimization is lost), and the idea is to come up with approach to eliminate this overhead and weakness. In this case, it is also important what memory exactly is being used, because trying not to use used parts of the same chunk also matters, because L-level caches of CPU might play a huge role in big data sets. One of the solutions is building quad-tree based data structure that will help optimize the calculation, and it has some expectation from data distribution which can speed up depending on dimensions (2D and 3D). This is a mutual problem of both data loading (memory usage) and its processing, and push-based system is a great approach, because it still remains as an abstract concept applicable to different queries and probably wide variety of other problems of similar kind in different fields.

\noindent\hspace{3em}Finally, we've analyzed some main ideas, advantages, and disadvantages of our approach in this paper which is basically push-based system with optimization on data processing part by means of GPU computation (in other words, parallelism). Reading the data from files is a huge overhead, thus, speedup of push-based system in general is significant over pull-based system in our given case. 
It's been demonstrated in the previous paper with different amount of atoms, frames, and workloads. Since we have to load the data every time there is a different query versus push-based system loads a chunk of data once, it is quite noticeable how the framework contributes to efficiency of analysis of simulation data. Even though the memory loading and pushing steps are the key features of the proposed design, nevertheless, having general knowledge about the network of queries, in this paper, we will try to focus on optimizing the actual execution of them.

\noindent\hspace{3em}Some of the queries may be very slow due to their nature on a sequential type of computation,  especially, 2-body functions. Since the nature of the data that comes with simulation is basically physical properties of atoms, there is a lot of computation that involves independent primitive mathematical operations, which makes it a perfect problem for parallelism. For example, image processing computations and graphics processing are done on GPUs, because they have a lot of minor primitive independent pixel-based computations.

\noindent\hspace{3em}As it has already been mentioned, although the original idea of push-based system for molecular simulation data analysis was focused on optimizing throughput and usage of loaded data, there were some extra approaches specifically for 2-body functions. For example, Density Map for Spacial Distance Histogram was used in order to avoid additional memory allocation and latency reduction with proven error bounds. SDH is a quite intensive and computational problem, especially with increasing amount of atoms.

\noindent\hspace{3em}We believe that having this nature of computational problems, the proposed GPU improved version of push-based system will significantly change in terms of performance by incorporating parallelism with CUDA.

@@

\section{Contribution And Roadmap Of The Paper}

@@

\chapter{RELATED WORK}\label{sc:relatedwork}

@@

\section{Scientific Software}

@@

\noindent\hspace{3em}The queries that we consider are related to computations limited within one frame of a molecular simulation. In general, with the given queries we have divided them into two categories: one body functions and multiple body functions. The first ones apparently have linear complexity of $O(n)$, while the other ones have bigger complexity.

\section{One Body Functions}
\hspace{3em} \emph{\textbf{One body functions}} such as sum of masses, center of mass, or simple counting with selection  are of a complexity $O(n)$, thus, in order to get the results for them going through every atom (essentially over the entire data) only once is enough, and taking into consideration the fact that we are going with push-based system, it is done at the same time for all of them.
